Wide binaries in Gaia DR1

### 2016-09-28 paper/ms.tex
partitioned matrix

% define macros for text
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\rave}{\project{\acronym{RAVE}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\figname}{Figure}

@@

For a pair of stars $(i,j)$, we want to ratio the fully marginalized likelihood
(FML) of the stars having the same 3-space velocity (hypothesis 1,
$\mathcal{L}_1$) over the FML of the stars having different 3-space velocities
(hypothesis 2, $\mathcal{L}_2$). When this ratio is large, the pair is a
candidate binary star. For \gaia\ DR1, we only observe parallax, $\varpi$, and
two proper motion components, $\propm = (\mu_\alpha, \mu_\delta)$ for each star.
Where there is overlap with \rave\ or \apogee, some number of observed radial
velocities, $v_r$, with observed variances $\sigma^2_{v_r}$ (independent of the
\gaia\ data) may also be available. We assume that the uncertainties in these
observed quantites are known with covariance matrix $\matrx{C}$. The likelihoods
$\mathcal{L}_1, \mathcal{L}_2$ are marginalized over true 3-space velocity and
distance for each star in the pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i) \, p(d_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i) \, p(d_j) \\
\end{align}
where
\begin{align}
  L(\bs{v}, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{\left(\varpi, \mu_\alpha, \mu_\delta, v_r \right)} \\
  \bs{x}_\theta &= \transp{\left(\frac{1}{d}, \frac{v_\alpha}{d},
    \frac{v_\delta}{d}, \tilde{v}_r \right)}
\end{align}
We assume Gaussian priors on the true velocity components and use a \todo{which
one will we use?} distance prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{align}
  p(\bs{v}) &= \mathcal{N}(0,25)~\kms \\
  p(d) &= \frac{r^2}{2\,L^3} \, \exp\left(-r/L\right)
\end{align}

@@

The integrals for Hypothesis 2 are somewhat easier because the true velocity
vector can be considered to be already in the orthogonal basis defined by the
tangent plane at the coordinates of the source (we are assuming an isotropic
velocity prior). We will analytically integrate over the velocity components
$(v_\alpha,v_\delta,v_r)$ and leave the distance integral to be done
numerically:
\begin{equation}
  \int \, \dd^3\bs{v} \, L(\bs{v}, d) \, p(\bs{v})
\end{equation}
The covariance matrix over parallax, proper motion components, and
radial velocity (if they exist) is $\matrx{C}$ and the inverse variance matrix
is $\matrx{V}$. Defining the constant
\begin{equation}
  \mathcal{Z} = \left[ \det \left(\frac{\matrx{V}}{2\pi} \right) \right]^{1/2}
\end{equation}
the integrand can be written
\begin{equation}
  L(\bs{v}, d) \, p(\bs{v}) = \mathcal{Z} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{V} \,
    \left(\bs{x} - \bs{x}_\theta \right)
    - \frac{|\bs{v}|^2}{2\,\sigma^2} \right] \\
\end{equation}

>>>>

% define macros for text
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\rave}{\project{\acronym{RAVE}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\figname}{Figure}
\newcommand{\eqname}{Equation}

@@

For a pair of stars $(i,j)$, we want to ratio the fully marginalized likelihood
(FML) of the stars having the same 3-space velocity (hypothesis 1,
$\mathcal{L}_1$) over the FML of the stars having different 3-space velocities
(hypothesis 2, $\mathcal{L}_2$). When this ratio is large, the pair is a
candidate binary star. For \gaia\ DR1, we only observe parallax, $\varpi$, and
two proper motion components, $\propm = (\mu_\alpha, \mu_\delta)$ for each star.
Where there is overlap with \rave\ or \apogee, some number of observed radial
velocities, $v_r$, with observed variances $\sigma^2_{v_r}$ (independent of the
\gaia\ data) may also be available. We assume that the uncertainties in these
observed quantites are known with covariance matrix $\matrx{C}$. The likelihoods
$\mathcal{L}_1, \mathcal{L}_2$ are marginalized over true 3-space velocity and
distance for each star in the pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i) \, p(d_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i) \, p(d_j) \\
\end{align}
where
\begin{align}
  L(\bs{v}, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{\left(\begin{array}{c c c c} \varpi & \mu_\alpha &
    \mu_\delta & v_r \end{array}\right)} \\
  \bs{x}_\theta &= \transp{\left(\begin{array}{c c c c} \frac{1}{d} & \frac{v_\alpha}{d} &
    \frac{v_\delta}{d} & \tilde{v}_r \end{array}\right)}
\end{align}
\todo{Holy fuck that vector is UGLY}
We assume Gaussian priors on the true velocity components and use a \todo{which
one will we use?} distance prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{align}
  p(\bs{v}) &= \mathcal{N}(0,25)~\kms \\
  p(d) &= \frac{r^2}{2\,L^3} \, \exp\left(-r/L\right)
\end{align}

@@

The integrals for Hypothesis 2 are somewhat easier because the true velocity
vector can be considered to be already in the orthogonal basis defined by the
tangent plane at the coordinates of the source (we are assuming an isotropic
velocity prior). We will analytically integrate over the velocity components
$(v_\alpha,v_\delta,v_r)$ and leave the distance integral to be done
numerically:
\begin{equation}
  \int \, \dd^3\bs{v} \, L(\bs{v}, d) \, p(\bs{v})
\end{equation}
The covariance matrix over parallax, proper motion components, and
radial velocity (if they exist) is $\matrx{C}$ and the inverse variance matrix
is $\matrx{V}$. Defining the constant
\begin{equation}
  \mathcal{Z} = \left[ \det \left(\frac{\matrx{V}}{2\pi} \right) \right]^{1/2}
\end{equation}
the integrand can be written
\begin{equation}
  L(\bs{v}, d) \, p(\bs{v}) = \mathcal{Z} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{V} \,
    \left(\bs{x} - \bs{x}_\theta \right)
    - \frac{|\bs{v}|^2}{2\,\sigma^2} \right] \\
\end{equation}
The integral over radial velocity is simple and can be performed without
considering any other terms because uncertainties in these measurements will
not, in general, be covariant with the Gaia data. The integrand over tangential
velocities can then be written
\begin{equation}
  L(v_\alpha, v_\delta, d) \, p(v_\alpha, v_\delta) = \tilde{\mathcal{Z}} \,
    \exp \left[ -\frac{1}{2}\,\transp{\left(\bs{x}' - \bs{x}'_\theta \right)} \,
    \matrx{V}' \,
    \left(\bs{x}' - \bs{x}'_\theta \right)
    - \frac{(v_\alpha^2 + v_\delta^2)}{2\,\sigma^2} \right]
    \label{eq:vtan-integrand}
\end{equation}
where the primed variables indicate that they are in the sub-space of just the
astrometric quantities, the constant pre-factor now includes the integral over
radial velocity
\begin{align}
  \tilde{\mathcal{Z}} = \mathcal{Z} \,
    \exp\left[-\frac{1}{2}\frac{v_r^2}{\sigma_{v_r}^2 + \sigma^2}\right] \,
    \left[ 2\pi \, (\sigma_{v_r}^2 + \sigma^2) \right]^{1/2}
\end{align}
and $\sigma_{v_r}$ is the observational uncertainty of the radial velocity
measurement. To perform the integral over tangential velocities, we partition
the inverse variance matrix and vectors so that
\begin{align}
  \bs{x}_A &= (\varpi)\\
  \bs{x}_B &= (\mu_\alpha, \mu_\delta)\\
  \bs{x}_{\theta, A} &= (d^{-1}) \\
  \bs{x}_{\theta, B} &= (d^{-1}\,v_\alpha, d^{-1}\,v_\delta) \\
  \matrx{V} &=
    \left(\begin{array}{c| c c}
      V_{11} & V_{12} & V_{13}\\\hline
      V_{12} & V_{22} & V_{23}\\
      V_{13} & V_{23} & V_{33}
    \end{array}\right)
    = \left(\begin{array}{c c}
      \matrx{V}_{AA} & \matrx{V}_{AB}\\
      \matrx{V}_{AB} & \matrx{V}_{BB}
    \end{array}\right)
\end{align}
\todo{Holy fuck that matrix is UGLY - fix with new version of AASTex 6}
and the argument of the Gaussian in \eqname~\ref{eq:vtan-integrand} can be
written
\begin{multline}
  -\frac{1}{2}\,[
  \transp{(\bs{x}_A - \bs{x}_{\theta, A})}\,
    \matrx{V}_{AA}\,(\bs{x}_A - \bs{x}_{\theta, A})\\
  + \transp{(\bs{x}_B - \bs{x}_{\theta, B})}\,
    \matrx{V}_{BB}\,(\bs{x}_B - \bs{x}_{\theta, B})\\
  + 2\,\transp{(\bs{x}_A - \bs{x}_{\theta, A})}\,
    \matrx{V}_{AB}\,(\bs{x}_B - \bs{x}_{\theta, B})]
\end{multline}
\todo{prior in above}
